---
title: "Climate-driven elevational range dynamics of dominant trees in the Western Himalayas"

author:
  - name: Abhishek Kumar
    url: https://akumar.netlify.app/
    orcid: 0000-0003-2252-7623
    email: abhikumar.pu@gmail.com
    affiliation: Panjab University, Chandigarh
    
  - name: Meenu Patil
    url: https://www.researchgate.net/profile/Meenu-Patil
    orcid: 0000-0002-7664-7877
    email: patilmeenu2@gmail.com
    affiliation: Panjab University, Chandigarh
    
  - name: Pardeep Kumar
    url: https://www.researchgate.net/profile/Pardeep-Kumar-22
    orcid: 0000-0001-6707-1485
    email: pardeepmor989@gmail.com
    affiliation: Panjab University, Chandigarh
    
  - name: Anand Narain Singh
    url: https://www.researchgate.net/profile/Anand-Singh-15
    orcid: 0000-0002-0148-8680
    email: dranand1212@gmail.com
    affiliation: Panjab University, Chandigarh

bibliography: refs.bib
csl: ecological-applications.csl

format: 
  # html:
  #   number-sections: true
  #   toc: true
  #   toc-depth: 3
  #   default-image-extension: png
  
  pdf:
    documentclass: scrartcl
    papersize: letter
    number-sections: true
    mainfont: Arial
    fontsize: 10pt
    linestretch: 1.5
    sansfont: Arial
    colorlinks: true
    link-citations: true
    citecolor: "#0000ff"
    default-image-extension: pdf
    code-block-bg: true
  
  docx:
    reference-doc: ecol-app.docx
    link-citations: true
    link-bibliography: true
    fig-dpi: 600
    default-image-extension: png
    
crossref: 
  title-delim: "  "
  fig-title:  "FIGURE S"  ## caption title
  fig-prefix: "Figure S"  ## inline title
  tbl-title:  "TABLE S"
  tbl-prefix: "Table S"
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  comment = "#>", echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE
)

## load packages
library(biomod2)        ## species distribution modelling
library(emmeans)        ## estimate marginal means
library(sf)             ## vector data handling
library(terra)          ## raster data handling
library(tidyverse)      ## general data manipulation and visualisation
library(tmap)           ## visualising spatial data

## colour palette for study sites
scen_pal <- c("current" = "#440154ff", "2040" = "#31688eff", 
              "2070" = "#35b779ff", "2100" = "#fde725ff")

## set theme for ggplot
theme_set(
  theme_bw(base_size = 14) +
    
    theme(
      panel.grid = element_blank(),
      
      axis.text = element_text(size = 10),
      
      strip.background = element_blank(),
      strip.text = element_text(hjust = 0, face = "bold")
    )
)
```

**Affiliation:** Soil Ecosystem and Restoration Ecology Lab, Department of Botany, Panjab University, Chandigarh 160014, INDIA

**Correspondence:** Anand Narain Singh. Email: <dranand1212@gmail.com>; <ansingh@pu.ac.in>

**Contact author:** Abhishek Kumar. Email: <abhikumar.pu@gmail.com>

**Co-first authors:** Abhishek Kumar and Meenu Patil are equal contributors to this work and designated as co-first authors.

**Open Research Statement:** Data and `R` codes are provided as private-for-peer review via the following link: <https://github.com/kumar-a/shifting-distributions>. Upon acceptance data will be archived in FigShare or Zenodo.

\newpage
# S1 Supplementary tables

```{r}
#| label: tbl-occ-literature
#| tbl-cap: Summary of literature sources used to obtain occurrence data for selected tree species in the Western Himalayas. 


read.csv("output/cleaned_occ.csv") |>
  
  ## unique species and reference
  distinct(species, reference) |>
  
  ## omit Field data
  filter(reference != "Field") |> 
  
  ## prepare species names
  mutate(species = paste0("*", species, "*")) |>
  
  ## prepare references
  mutate(refs = paste0("@", reference)) |>
  arrange(reference) |>
  
  ## make wider format
  pivot_wider(names_from = reference, values_from = refs, values_fill = "-") |>
  arrange(species) |>
  unite(col = "Reference", Amare2016:Tiwari2023, sep = "; ") |>
  mutate(Reference = paste0("[", gsub("-; |; -", "", Reference), "]")) |>
  
  ## make table
  rename("Species" = species) |>
  knitr::kable()
```

\newpage

```{r}
#| label: tbl-net-change
#| tbl-cap: Predicted net change in habitat suitability for selected species under future (2040, 2070, and 2100) climatic conditions. The values represent the percentage of pixels in the study area, and species are arranged according to net change by the end of the century (2100). 


list.files(
  path = "output/range_dynamics",
  pattern = "src_count.csv",
  full.names = TRUE
) |> 
  read_csv(id = "path") |>
  mutate(path = gsub("output/range_dynamics/|_current_to|_src_count.csv", "", path)) |>
  separate(path, into = c("species", "year"), sep = "_") |>
  filter(grepl("EMca", full.name)) |>
  mutate(Species = paste0("*", gsub("\\.", " ", species), "*") ) |>
  select(Species, year, perc.level, perc.value) |>
  distinct() |>
  pivot_wider(names_from = perc.level, values_from = perc.value) |>
  select(Species, year, SpeciesRangeChange) |>
  pivot_wider(names_from = year, values_from = SpeciesRangeChange) |>
  arrange(`2100`) |>
  knitr::kable(digits = 2)
```

\newpage

```{r}
#| label: tbl-elev-anova
#| tbl-cap: Effects of climatic scenarios (current, 2040, 2070, and 2100) on elevational distribution for each species. The type-II ANOVA was used to determine the significance of effects, and generalized eta squared (&eta;^2^) was used as effect size.


read_csv("output/suitability_elevation.csv", show_col_types = FALSE) |>
  dplyr::group_by(species) |>
  rstatix::anova_test(elevation ~ scen, type = 2) |>
  as_tibble() |>
  mutate(species = paste0("*", gsub("\\.", " ", species), "*"),
         p = ifelse(p < 0.001, yes = "<0.001", as.character(round(p, 3))),
         `F` = round(`F`, 2)
         ) |>
  select(-c(Effect, `p<.05`)) |>
  rename("Species" = species, "&eta;^2^" = ges) |>
  knitr::kable(align = "lrrrrr")
```

\newpage

```{r}
#| label: tbl-elev-comp
#| tbl-cap: Comparison of average elevation under current and future climatic conditions for selected species in the Western Himalayas. The values represent the estimated marginal mean elevation (in meters) and associated standard error (SE) across all predicted pixels. The significance of difference was determined using the t-test with Tukey's adjustment for multiple comparisons (equivalent to Tukey's test). Within each row, the values superscripted with shared letters indicate non-significant differences (p > 0.05).


read_csv("output/suitability_elevation.csv", show_col_types = FALSE) |>
  dplyr::group_by(species) |>
  tidyr::nest() |>
  rowwise() |> 
  dplyr::mutate(aov_results = list(lm(elevation ~ scen, data = data)), 
         emm = list(emmeans::emmeans(aov_results, "scen", type= "response")), 
         cld = list(multcomp::cld(emm, Letters = letters, reverse = FALSE))) |> 
  dplyr::select(-data, -aov_results, -emm)|> 
  unnest(cols = c(cld)) |>
  dplyr::mutate(cld = trimws(.group)) |> 
  dplyr::select(species, scen, emmean, SE, cld) |>
  mutate(across(is.double, ~round(.x, 2))) |>
  mutate(species = paste0("*", gsub("\\.", " ", species), "*"),
         Elevation = paste0(emmean, " ± ", SE, "^", cld, "^")
         ) |>
  dplyr::select(species, scen, Elevation) |>
  pivot_wider(names_from = scen, values_from = Elevation) |>
  select(species, current, `2040`, `2070`, `2100`) |>
  rename("Species" = species, "Current" = current) |>
  knitr::kable(align = "lrrrr")
```

\newpage

```{r}
#| label: tbl-range-param-anova
#| tbl-cap: Effects of climatic scenarios (current, 2040, 2070, and 2100) on elevation range parameters. The type-II ANOVA was used to test the significance of effects, and generalized eta squared (&eta;^2^) was used as effect size.


## optimum elevation
elev_opt <- read_csv("output/suitability_elevation.csv", show_col_types = FALSE) |>
  ## build 100 m elevational bands
  mutate(elev_band  = cut(elevation, breaks = c(seq(0, 4900, 100), Inf),
                          labels = c(seq(100, 5000, 100)))) |>
  ## convert to numeric
  mutate(elev_band = as.character(elev_band),
         elev_band = as.numeric(elev_band)) |>
  ## count number of pixels
  group_by(species, scen) |>
  count(elev_band) |>
  ## filter elevational band with maximum pixels
  group_by(species, scen) |>
  filter(n == max(n)) |>
  select(-n) |>
  rename("Optimum.range" = elev_band)

## elevation range parameters (optimum and edges) ANOVA

read_csv("output/suitability_elevation.csv", show_col_types = FALSE) |>
  left_join(elev_opt, by = join_by(species, scen)) |>
  group_by(species, scen) |>
  mutate(range_params = case_when(
    elevation < quantile(elevation, 0.05) ~"Trailing",
    between(elevation, Optimum.range - 100, Optimum.range) ~"Optimum",
    elevation > quantile(elevation, 0.95) ~"Leading",
    TRUE ~ "Other"
  )) |>
  select(-Optimum.range) |>
  filter(range_params != "Other") |>
  group_by(species, range_params) |>
  rstatix::anova_test(formula = elevation ~ scen, type = 2) |>
  as_tibble() |>
  mutate(
    species = paste0("*", gsub("\\.", " ", species), "*"),
    `F` = round(`F`, 2),
    p = ifelse(p < 0.001, yes = "<0.001", as.character(round(p, 3))),
    ges = ifelse(ges < 0.001, yes = "<0.001", as.character(round(ges, 3)))
  ) |>
  select(-c(Effect, `p<.05`)) |>
  rename("Species" = species, "Parameters" = range_params, "&eta;^2^" = ges) |>
  knitr::kable(align = "llrrrrr")
```

\newpage

```{r}
## optimum elevation
elev_opt <- read_csv("output/suitability_elevation.csv", show_col_types = FALSE) |>
  ## build 100 m elevational bands
  mutate(elev_band  = cut(elevation, breaks = c(seq(0, 4900, 100), Inf),
                          labels = c(seq(100, 5000, 100)))) |>
  ## convert to numeric
  mutate(elev_band = as.character(elev_band),
         elev_band = as.numeric(elev_band)) |>
  ## count number of pixels
  group_by(species, scen) |>
  count(elev_band) |>
  ## filter elevational band with maximum pixels
  group_by(species, scen) |>
  filter(n == max(n)) |>
  select(-n) |>
  rename("Optimum.range" = elev_band)

## elevation range parameters (optimum and edges)
elev_range_param <- read_csv("output/suitability_elevation.csv", 
                             show_col_types = FALSE) |>
  left_join(elev_opt, by = join_by(species, scen)) |>
  group_by(species, scen) |>
  mutate(range_params = case_when(
    elevation < quantile(elevation, 0.05) ~"Trailing.edge",
    between(elevation, Optimum.range - 100, Optimum.range) ~"Optimum.range",
    elevation > quantile(elevation, 0.95) ~"Leading.edge",
    TRUE ~ "Other.range"
  )) |>
  select(-Optimum.range) |>
  filter(range_params != "Other.range")

## tukey's test for significance
tuk_trailing <- elev_range_param |>
  filter(range_params == "Trailing.edge") |>
  dplyr::group_by(species) |>
  tidyr::nest() |>
  rowwise() |> 
  dplyr::mutate(aov_results = list(lm(elevation ~ scen, data = data)), 
                emm = list(emmeans::emmeans(aov_results, "scen", type = "response")), 
                cld = list(multcomp::cld(emm, Letters = letters, reverse = FALSE))) |> 
  dplyr::select(-data, -aov_results, -emm)|> 
  unnest(cols = c(cld)) |>
  dplyr::mutate(cld = trimws(.group)) |> 
  dplyr::select(-.group) |>
  select(species, scen, emmean, SE, cld)

## tukeys test for optimum range
tuk_optimum <- elev_range_param |>
  filter(range_params == "Optimum.range") |>
  dplyr::group_by(species) |>
  tidyr::nest() |>
  rowwise() |> 
  dplyr::mutate(aov_results = list(lm(elevation ~ scen, data = data)), 
                emm = list(emmeans::emmeans(aov_results, "scen", type= "response")), 
                cld = list(multcomp::cld(emm, Letters = letters, reverse = FALSE))) |> 
  dplyr::select(-data, -aov_results, -emm)|> 
  unnest(cols = c(cld)) |>
  dplyr::mutate(cld = trimws(.group)) |> 
  dplyr::select(-.group) |>
  select(species, scen, emmean, SE, cld)

## tukeys test for leading edge
tuk_leading <- elev_range_param |>
  filter(range_params == "Leading.edge") |>
  dplyr::group_by(species) |>
  tidyr::nest() |>
  rowwise() |> 
  dplyr::mutate(aov_results = list(lm(elevation ~ scen, data = data)), 
                emm = list(emmeans::emmeans(aov_results, "scen", type= "response")), 
                cld = list(multcomp::cld(emm, Letters = letters, reverse = FALSE))) |> 
  dplyr::select(-data, -aov_results, -emm)|> 
  unnest(cols = c(cld)) |>
  dplyr::mutate(cld = trimws(.group)) |> 
  dplyr::select(-.group) |>
  select(species, scen, emmean, SE, cld)
```

```{r}
#| label: tbl-range-param-comp
#| tbl-cap: Comparison of elevational range parameters under current and future climatic conditions for selected species in the Western Himalayas. The values represent the estimated marginal mean elevation (in meters) and associated standard error (SE). The significance of difference was determined using the t-test with Tukey's adjustment for multiple comparisons (equivalent to Tukey's test). Within each row, the values superscripted with shared letters indicate non-significant differences (p > 0.05).


bind_rows(
  mutate(tuk_trailing, Parameters = "Trailing"),
  mutate(tuk_optimum,  Parameters = "Optimum"),
  mutate(tuk_leading,  Parameters = "Leading")
) |>
  ungroup() |>
  mutate(across(is.double, ~round(.x, 2))) |>
  mutate(Species = paste0("*", gsub("\\.", " ", species), "*"),
         Elevation = paste0(emmean, " ± ", SE, "^", cld, "^")
         ) |>
  dplyr::select(Species, Parameters, scen, Elevation) |>
  pivot_wider(names_from = scen, values_from = Elevation) |>
  arrange(Species) |>
  rename("Current" = current) |>
  knitr::kable(align = "llrrrr")
```

\newpage

```{r}
#| label: tbl-slope-values
#| tbl-cap: Elevational changes in species range parameters due to current (2010) and future (2040, 2070, and 2100) climatic conditions. The values are estimated slope values (Estimate) with associated standard errors (SE) from linear regression of marginal means against the years representing climatic conditions.

df_trailing <- tuk_trailing |> 
  select(species, scen, emmean) |>
  pivot_wider(names_from = scen, values_from = emmean) |>
  mutate(Parameters = "Trailing")

df_optimum <- tuk_optimum |> 
  select(species, scen, emmean) |>
  pivot_wider(names_from = scen, values_from = emmean) |>
  mutate(Parameters = "Optimum")

df_leading <- tuk_leading |> 
  select(species, scen, emmean) |>
  pivot_wider(names_from = scen, values_from = emmean) |>
  mutate(Parameters = "Leading")

bind_rows(df_trailing, df_optimum, df_leading) |> 
  pivot_longer(cols = c(current, `2040`, `2070`, `2100`), 
               names_to = "scen", values_to = "Elevation") |>
  mutate(scen = ifelse(scen == "current", "2010", scen),
         scen = as.numeric(scen)) |>
  
  group_by(species, Parameters) |>
  nest() |>
  mutate(Model = map(data, ~lm(Elevation ~ scen, data = .x) |> broom::tidy())
         ) |> 
  select(-data) |>
  unnest(Model) |>
  filter(term == "scen") |>
  mutate(p.value = round(p.value, 3),
         p.value = ifelse(p.value < 0.05, 
                          yes = paste0("**", p.value, "**"), 
                          no = as.character(p.value))) |>
  mutate(species = paste0("*", gsub("\\.", " ", species), "*"),
         Parameters = factor(Parameters, c("Trailing", "Optimum", "Leading"))) |>
  select(-term) |>
  mutate(across(is.double & !p.value, ~round(.x, 2))) |>
  arrange(species) |>
  knitr::kable(col.names = c("Species", "Parameters", "Estimate", "SE", "z", "p"),
               align = "llrrrr")
```

```{r}
# bind_rows(
#   mutate(tuk_trailing, Parameters = "Trailing"),
#   mutate(tuk_optimum,  Parameters = "Optimum"),
#   mutate(tuk_leading,  Parameters = "Leading")
# ) |>
#   ungroup() |>
#   mutate(across(is.double, ~round(.x, 2))) |>
#   select(-c(SE, cld)) |>
#   pivot_wider(names_from = scen, values_from = emmean) |>
#   mutate(Shift = `2100` - current) |>
#   ggplot(aes(x = Parameters, y = species, fill = Shift)) +
#   geom_tile(color = "white") +
#   geom_text(aes(label = round(Shift, 2))) +
#   scale_fill_steps2(n.break = 7)
```

\newpage
# S2 Supplementary figures

```{r}
# DiagrammeR::grViz("clean_occ.gv") |>
#   DiagrammeRsvg::export_svg() |> 
#   charToRaw() |> 
#   rsvg::rsvg_pdf("figs/s01-occ_cleaning.pdf")
#   # rsvg::rsvg_png("figs/s01-occ_cleaning.png")
```

![Flow diagram showing the pre-processing and cleaning of geographical occurrences for selected species.](figs/s01-occ_cleaning){#fig-occ-cleaning width=70%}

```{r fig.width=7, fig.height=5}
#| label: fig-species-occurrence
#| fig-cap: Geographic distribution of cleaned occurrence dataset for selected species in the Western Himalayas.


## Western Himalayan States
wh <- st_read("data/site_states.gpkg", quiet = TRUE) |>
  filter(STATE %in% c("Himachal Pradesh", "Uttarakhand"))

read.csv("output/cleaned_occ.csv") |>
  # count(species, sort = TRUE)
  st_as_sf(coords = c("longitude", "latitude"), crs = "epsg:4326") |>
  
  ggplot() +
  geom_sf(data = wh) +
  geom_sf(color = "red", alpha = 0.25) +
  facet_wrap(.~species, nrow = 3) +
  theme_void() +
  theme(strip.text = element_text(size = 9, hjust = 0, face = "italic"))

## save the plots
# ggsave("figs/s02-spec_occ.pdf", width = 7, height = 5)
# ggsave("figs/s02-spec_occ.png", width = 7, height = 5, dpi = 600)
```



```{r fig.width=8, fig.height=6}
#| label: fig-bio-current
#| fig-cap: Bioclimatic variables used for modeling the potential habitat suitability (a--f). The mean diurnal air temperature range (bio2), isothermality (bio3), mean daily air temperatures of the wettest quarter (bio8), precipitation amount of the driest month (bio14), precipitation seasonality (bio15) and mean monthly precipitation amount of the warmest quarter (bio18) were downloaded from CHELSA database.

## function to plot bio
bio.plot <- function(bioname){
  
  ## get the bionumber
  bionumber <- gsub("bio", "", bioname) |> as.numeric()
  
  ## Western Himalayan States
  wh <- st_read("data/site_states.gpkg", quiet = TRUE) |>
    filter(STATE %in% c("Himachal Pradesh", "Uttarakhand"))
  
  ## format bio data
  bio <- rast("data/wh_bio_2010_current.tif", lyr = bionumber) |>
    as.data.frame(xy = TRUE) |>
    rename(bioname = 3)
  
  ## plot the map
  ggplot() +
    geom_raster(data = bio, aes(x, y, fill = bio[,3])) +
    geom_sf(data = wh, fill = NA) +
    theme_void() +
    theme(legend.position = "top")
}

pb2 <- bio.plot("bio2") +
  scale_fill_viridis_b(
    option = "rocket", direction = 1, 
    guide = guide_coloursteps(barwidth = 8, barheight = 0.75,
                              label.position = "top",
                              title = "bio2",
                              title.vjust = 0.1)
  )

pb3 <- bio.plot("bio3") +
  scale_fill_fermenter(
    palette = "YlOrBr", direction = -1, n.breaks = 4,
    guide = guide_coloursteps(barwidth = 8, barheight = 0.75,
                              label.position = "top",
                              title = "bio3",
                              title.vjust = 0.1)
  )

pb8 <- bio.plot("bio8") +
  scale_fill_fermenter(
    palette = "PuOr", direction = -1, n.breaks = 4,
    guide = guide_coloursteps(barwidth = 8, barheight = 0.75,
                              label.position = "top",
                              title = "bio8",
                              title.vjust = 0.1)
  )

pb14 <- bio.plot("bio14") +
  scale_fill_fermenter(
    palette = "Blues", direction = 1,
    guide = guide_coloursteps(barwidth = 8, barheight = 0.75, 
                              label.position = "top",
                              title = "bio14",
                              title.vjust = 0.1)
  )

pb15 <- bio.plot("bio15") +
  scale_fill_fermenter(
    palette = "YlGnBu", direction = 1,
    guide = guide_coloursteps(barwidth = 8, barheight = 0.75, 
                              label.position = "top",
                              title = "bio15",
                              title.vjust = 0.1)
  )

pb18 <- bio.plot("bio18") +
  scale_fill_fermenter(
    palette = "Purples", direction = 1,
    guide = guide_coloursteps(barwidth = 8, barheight = 0.75, 
                              label.position = "top",
                              title = "bio18",
                              title.vjust = 0.1)
  )

## arrange plots
cowplot::plot_grid(
  pb2, pb3, pb8, pb14, pb15, pb18,
  nrow = 2, labels = paste0(letters[1:6], ")"),
  label_y = 0.96
)

## save the plots
# ggsave("figs/s03-bio_current.pdf", width = 8, height = 6)
# ggsave("figs/s03-bio_current.png", width = 8, height = 6, dpi = 600)
```


```{r fig.width=7, fig.height=6}
#| label: fig-mp-bg-opt
#| fig-cap: Effect of sampling method (disk vs random) and the number of background points on the performance of MaxEnt models for *Mallotus philippensis* in the Western Himalayas. The disk method sampled random points from the background defined with a minimum of 1 km and a maximum of 50 km to the presence occurrence points. The initial dataset was randomly divided into 70% for training (Calibration) and 30% for testing (Validation). Each model was repeated three times for cross-validation. The performance of the model was assessed using the true skill statistic (TSS) and the area under the receiver operating characteristic curve (ROC). The values represent the mean and SD of three replications. The shaded grey region indicates the finally selected number of background points (n &equals; 1000).


read.csv("output/mallotus_background.csv") |>
  pivot_longer(cols = c(calibration, validation),
               names_to = "dataset", values_to = "y") |>
  select(PA, y, dataset, metric.eval, pa.meth) |>
  mutate(dataset = str_to_title(dataset)) |>
  group_by(PA, dataset, metric.eval, pa.meth) |>
  rstatix::get_summary_stats("y", type = "mean_sd") |>

  ggplot(aes(x = PA, y = mean, ymin = mean - sd, ymax = mean + sd,
             fill = pa.meth, shape = pa.meth)) +
  geom_rect(aes(xmin = 2.5, xmax = 3.5, ymin = -Inf, ymax = Inf), 
            fill = "grey90", alpha = 0.2) +
  geom_pointrange(position = position_dodge(0.5), size = 0.75,
                  stroke = 0.5) +
  facet_grid(metric.eval ~ dataset) +
  scale_fill_manual(values = c("#e69f00", "#56b4e9"), name = "Method") +
  scale_shape_manual(values = c(21, 24), name = "Method") +
  scale_x_discrete(labels = c(250, 500, 1000, 2000, 5000)) +
  labs(x = "Background points", y = "") +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank(),
        panel.spacing = unit(5, "mm"),
        legend.title = element_blank(),
        legend.position = c(0.005, 0.525), 
        legend.justification = c(0, 0),
        legend.background = element_rect(colour = "grey", linewidth = 0.25),
        strip.background = element_rect(fill = "grey90"),
        strip.text = element_text(face = "bold"))

## save the plot
# ggsave("figs/s04-background_points.pdf", width = 7, height = 6)
# ggsave("figs/s04-background_points.png", width = 7, height = 6, dpi = 600)
```

```{r}
## plot continuous habitat suitability
# source("R/10-plot_hab_suit.R")
```

![Habitat suitability for selected species under current and future (2040, 2070, and 2100) climatic conditions over the Western Himalayas. The predictions indicate the ensemble model of all individual models with a TSS greater than 0.7 using the weighted mean averaging (EMwmean).](figs/s05-hab_suit){#fig-hab-suit}

```{r fig.width=7, fig.height=5}
#| label: fig-elev-change
#| fig-cap: Elevational distribution of pixels indicating (a) gain, (b) stable, and (c) loss in assemblage level habitat suitability (species richness) under future climatic conditions.

## prepare the elevation change data
elev_change <- read.csv("output/richness_elevation.csv") |>
  select(scen, richness, elevation) |>
  mutate(richness = case_when(richness <  0  ~"c) Loss",
                              richness == 0  ~"b) Stable",
                              richness >  0  ~"a) Gain"),
         elevation = cut(elevation, seq(0, 5000, 100), seq(100, 5000, 100)),
         elevation = as.numeric(as.character(elevation))/1e3) |>
  group_by(richness) |>
  count(elevation) |>
  mutate(perc_n = n/sum(n)*100)

## plot pixels distribution for richness gains
p_gain <- elev_change |>
  filter(richness == "a) Gain") |>
  ggplot(aes(x = elevation, y = perc_n)) +
  geom_bar(stat = "identity", fill = "#5ab4ac", width = 0.075) +
  xlim(0, 5) +
  theme(axis.title = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

## plot pixels distribution for stable richness
p_stable <- elev_change |>
  filter(richness == "b) Stable") |>
  ggplot(aes(x = elevation, y = perc_n)) +
  geom_bar(stat = "identity", fill = "grey", width = 0.075) +
  xlim(0, 5) +
  labs(y = "Pixels (%)") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

## plot pixels distribution for richness loss
p_loss <- elev_change |>
  filter(richness == "c) Loss") |>
  ggplot(aes(x = elevation, y = perc_n)) +
  geom_bar(stat = "identity", fill = "#d8b365", width = 0.075) +
  xlim(0, 5) +
  labs(x = "Elevation (km)") +
  theme(axis.title.y = element_blank())

## arrange the plots
cowplot::plot_grid(
  p_gain, p_stable, p_loss, 
  ncol = 1, align = "v",
  labels = paste0(letters[1:3], ")"), 
  label_x = 0.075, label_y = 0.97,
  rel_heights = c(0.3, 0.3, 0.4)
)

## save the plot
# ggsave("figs/s06-elev_rich_change.pdf", width = 7, height = 6)
# ggsave("figs/s06-elev_rich_change.png", width = 7, height = 6, dpi = 600)
```

```{r}
## plot single response curves
# source("R/11-plot_resp_curve_single.R")
```

![Plot of the response curves of six environmental variables used to model habitat suitability of selected species. Model predictions were plotted against each environmental variable while holding the values of all other variables constant at their respective means. The grey lines represent the individual response curves from the evaluated 14 models (12 individual + 2 ensemble models). The smooth line (thick and colored) was fitted using the generalized additive model (GAM) with a shrinkage version of the penalized cubic spline. Temperature-related variables are shown with solid red lines and precipitation-related variables with solid blue lines. The right y-axis represents the probability of a suitable habitat, and the x-axis represents the environmental predictors in their respective units.](figs/s07-resp_curve_single){#fig-resp-curve-single width=70%}

\newpage
# S3 ODMAP protocol

```{r}
#| tbl-colwidths: [25, 75]

read.csv("odmap_protocol.csv") |>
  mutate(Description = ifelse(
    element == "", "",
    paste0("**", element, "**: ", Value)
  )) |>
  rename("Element" = subsection) |>
  select(Element, Description) |>
  knitr::kable(booktabs = TRUE, longtable = TRUE, linesep = "")

  # kableExtra::kbl(format = "markdown") |>
  # column_spec(1, width = "1.5in") |>
  # column_spec(2, width = "4.5in") |>
  # row_spec(c(1, 27, 61, 75, 79), background = "#d9e2f3") |>
  # collapse_rows(columns = 1, valign = "top")
```


```{r}
# library(kableExtra)
# 
# read.csv("odmap_protocol.csv") |>
#   mutate(Description = ifelse(
#     element == "", "",
#     paste0("**", element, "**: ", Value)
#   )) |>
#   rename("Element" = subsection) |>
#   select(Element, Description) |>
#   kbl(format = "markdown", booktabs = TRUE, longtable = TRUE) |>
#   column_spec(column = 1, width ="1.5in") |>
#   column_spec(column = 2, width ="4.5in") |>
#   row_spec(row = c(1, 27, 61, 75, 79), background = "#d9e2f3")
```


\newpage
# S4 Session Info

```{r}
sessioninfo::platform_info() |>
  as.data.frame() |>
  t() |>
  as.data.frame() |>
  rownames_to_column("setting") |>
  rename("value" = V1) |>
  knitr::kable()
```

```{r}
sessioninfo::package_info() |>
  as.tibble() |>
  select(-c(ondiskversion, path, loadedpath, library)) |>
  mutate(package = paste0("`", package, "`")) |>
  rename("version" = loadedversion) |>
  knitr::kable()
```


\newpage
# References {.unnumbered}
